---
title: "What is R Markdown?"
author: ""
date: ''
output:
  html_document:
      number_sections: false
---

**Welcome to R Markdown!**

# What is it about? 

In the world of reproducible research, you want your work and your analysis to be easily reproduced by other researchers. Otherwise, how am I going to believe that you have analyzed your data properly?\
\
**R Markdown allows you to create documents that serve as a neat record of your analysis**. With RMarkdown you can present your **code** alongside its output (graphs, tables, etc.) with conventional text to explain it, a bit like a notebook. You can also share your file by uploading it to an **online repository such as Github**.


## Create an R Markdown (.Rmd) file

* To get R Markdown working in RStudio, you need the `rmarkdown` package, which you can get from `CRAN` by running the following commands in R or RStudio:

```{r , include=TRUE, echo=TRUE}
#install.packages("rmarkdown")
#library(rmarkdown)
```

* Create a new RMarkdown file (.Rmd) by selecting: `File -> New File -> R Markdown`, then choose the file type you want to create.\


## YAML header material

The YAML header is the top section of an R Markdown script and it is enclosed by `---`. By default it includes:\

* Title\

* Author\

* Date\

* File Type (for the output)\
\

**IT IS TIME TO CREATE YOUR FIRST .Rmd FILE.** To compile your .Rmd file, you should press the `Knit` button in the taskbar.


## Markdown syntax

R Markdown makes use of [Markdown syntax](https://www.markdownguide.org/basic-syntax).


## Insert your code from an R script into a .Rmd file

Code chunks should be enclosed by three backwards apostrophes ` ``` `. Example of these code chunks include:

```{r}
sample <- rnorm(10, mean = 0, sd = 1)
```

* The `echo` option: this allows you to hide a particular code chunk from appearing in the final document, but still showing only the output of this chunk. Simply, set `echo = FALSE` in the code chunk instructions.\

* You might want to hide the code and the output in the final .html file. To do this you can use, `include = FALSE`. 

```{r , include=TRUE, echo=FALSE}
plot(cars)
```








